Exploring the Traveling Salesman Problem

### 2017-05-11 paper/paper.tex
typo

% Introduction
% ---
\section*{Introduction}
The Traveling Salesman Problem (TSP) is a classic NP---hard problem
in the field of computer science.  Like every other NP--hard problem,
it can only be solved in exponential runtime, or approximated in polynomail
runtime.  In this paper, we will obtain an understanding of the TSP problem
by explaining what it is, examining its input and output, and looking at
different algorithms that can obtain or approximate a solution.  We will
present and explain a naive brute force solution to the problem and analyze
its efficiency.  Then we will look at alternative TSP algorithms and discuess
the use cases and advantages of each approach.  We will choose one of the
algorithms and present its pseudocode, as well as detailed mathematical
notation to gain a better understanding of how it computes the solution.

@@

An example of a \textit{directed} variation of the TSP is one that
optimizes for minimum total airfair to travel a tour by plane,
where ticket prices from city $i$ to city $j$ is not necessarily
equal to the ticket price from city $j$ to city $i$, implying the
cost matrix is asymmetric.
\par

@@

TSP algorithms can also about a graph $G=(V,E)$, where $V_i$
represents city $i$ and $E_{ij}$ represents traveling from city $i$ to
city $j$ along the optimal tour, where the weight is the distance between
the two.  If the graph where visualized, each verticy would be positioned
at its longitude and latitude coordinates and a segment would connect each pair
of verticies $v_i$ and $v_j$ where $e_{ij}=1$.  The single path in the graph
represents the optimal tour.
\par

@@

\subsection{Held--Karp}
The Held--Karp algorithm is an $O(n^2 2^n)$ approach to solving the TSP\@.
It is not fast enough to solve TSP instances with a large number of
cities, but the solution for the instances it can solve is guarenteed
to be correct.  The algorithm works as follows:
\par

@@

\subsection{Christofides}
Christofide's algorithm is an $O(n^3)$ approximation algorithm that guarentees
the cost of its solution will be strictly within $3/2$ of the optimal solution
cost.  The algorithm only works on a symmetric distance matrix that satisfies
the triangle inequality \cite{nicos}.  That is, a distance matrix $D=d_{ij}$
may be used if the shortest path to reach every $j$ from
$i$ involves traveling directly to $j$ from $i$, rather than through a path that
includes more than two verticies.
\par

The algorithm works as follows: Given a distance matrix $D=d_{ij}$ and size $n$,
let $G=(V,E)$ where $v_i=i$ for all $i\in [1,\dots,n]$ and $e_{ij}=d_{ij}$.  The
optimal tour starting at $v_1$ is obtained using the following steps:
\begin{enumerate}
  \item Generate a minimum spanning tree, $MST$, rooted at $v_1$ using Prim's algorithm.
  \item Perform a preorder traversal on $MST$ and append the verticies in the order they 
    are visited to list a $T$.
  \item For each verticy in $T$, append it to $T'$ only when it is first visited.  This 
    removes duplicates from $T$ resulting in the optimal tour.
\end{enumerate}

@@

On lines 5-7, we add our base cases to the lookup tables.  In the main loop on
lines 9-25, we build the lookup tables in a bottom-up dynamic programming fashion.
Notice on lines 11 and 12 we continue when the subset size is $n$ and $x \neq 1$.
This acheives the effect of beginning our recurrence from (3.2) with $(\{1,\dots,n\}, 1)$,
ensuring we're only computing optimal paths that start at $1$.  Finally, on lines 27-31
we backtrack through our lookup table $P$ that stores the optimal $(S-\{x\}, k)$
for a given $(S, x)$, appending each optimal $k$ value to the optimal tour $T$.

@@

\begin{proof}
  (Induction) For our basis step, consider that an empty set $\varnothing$ has $1$
  subset (itself).  $2^n = 2^0 = 1$, so a set with $0$ elements has $2^0$ subsets.\
  For induction, suppose a $k$-element set has $2^k$ subsets.  We will show every
  $(k+1)$-element subset has $2^(k+1)$ subsets.  Let $A=\{a_1,a_2,\dots,a_k, b\}$,
  notice it contains $k+1$ elements.  We parition the subsets of $A$ into two sets
  $B$ and $C$, where $B$ contains the subsets that contain $b$ and $C$ contains
  the subsets that don't contain $b$.  $C$ contains every subset of $A-{b}$, so it
  has $2^k$ elements.  By construction, it follows that $B$ must have the same number
  of elements as $C$, so it also has $2^k$ elements.  Since $B$ and $C$ are independent
  sets, the size of $B\union C$ is $2^k + 2^k = 2^{k+1}$.
\end{proof}

@@

\newpage
\appendix
\section*{Appendices}
\addcontentsline{toc}{section}{Appendices}
\renewcommand{\thesubsection}{\Alph{subsection}}

\subsection{Collaboration}
  I collaborated with Grant Eaton and Luke Artnak.

\newpage
\subsection{Code Demonstration}
  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}

>>>>

% Introduction
% ---
\section*{Introduction}
The Traveling Salesman Problem (TSP) is a classic NP---hard problem
in the field of computer science.  Like every other NP--hard problem,
it can only be solved in exponential runtime, or approximated in polynomial
runtime.  In this paper, we will obtain an understanding of the TSP problem
by explaining what it is, examining its input and output, and looking at
different algorithms that can obtain or approximate a solution.  We will
present and explain a naive brute force solution to the problem and analyze
its efficiency.  Then we will look at alternative TSP algorithms and discuss
the use cases and advantages of each approach.  We will choose one of the
algorithms and present its pseudocode, as well as detailed mathematical
notation to gain a better understanding of how it computes the solution.

@@

An example of a \textit{directed} variation of the TSP is one that
optimizes for minimum total airfare to travel a tour by plane,
where ticket prices from city $i$ to city $j$ is not necessarily
equal to the ticket price from city $j$ to city $i$, implying the
cost matrix is asymmetric.
\par

@@

TSP algorithms can also output a graph $G=(V,E)$, where $V_i$
represents city $i$ and $E_{ij}$ represents traveling from city $i$ to
city $j$ along the optimal tour, where the weight is the distance between
the two.  If the graph where visualized, each vertex would be positioned
at its longitude and latitude coordinates and a segment would connect each pair
of vertices $v_i$ and $v_j$ where $e_{ij}=1$.  The single path in the graph
represents the optimal tour.
\par

@@

\subsection{Held--Karp}
The Held--Karp algorithm is an $O(n^2 2^n)$ approach to solving the TSP\@.
It is not fast enough to solve TSP instances with a large number of
cities, but the solution for the instances it can solve is guaranteed
to be correct.  The algorithm works as follows:
\par

@@

\subsection{Christofides}
Christofide's algorithm is an $O(n^3)$ approximation algorithm that guarantees
the cost of its solution will be strictly within $3/2$ of the optimal solution
cost.  The algorithm only works on a symmetric distance matrix that satisfies
the triangle inequality \cite{nicos}.  That is, a distance matrix $D=d_{ij}$
may be used if the shortest path to reach every $j$ from
$i$ involves traveling directly to $j$ from $i$, rather than through a path that
includes more than two vertices.
\par

The algorithm works as follows: Given a distance matrix $D=d_{ij}$ and size $n$,
let $G=(V,E)$ where $v_i=i$ for all $i\in [1,\dots,n]$ and $e_{ij}=d_{ij}$.  The
optimal tour starting at $v_1$ is obtained using the following steps:
\begin{enumerate}
  \item Generate a minimum spanning tree, $MST$, rooted at $v_1$ using Prim's algorithm.
  \item Perform a preorder traversal on $MST$ and append the vertices in the order they 
    are visited to list a $T$.
  \item For each vertex in $T$, append it to $T'$ only when it is first visited.  This 
    removes duplicates from $T$ resulting in the optimal tour.
\end{enumerate}

@@

On lines 5-7, we add our base cases to the lookup tables.  In the main loop on
lines 9-25, we build the lookup tables in a bottom-up dynamic programming fashion.
Notice on lines 11 and 12 we continue when the subset size is $n$ and $x \neq 1$.
This achieves the effect of beginning our recurrence from (3.2) with $(\{1,\dots,n\}, 1)$,
ensuring we're only computing optimal paths that start at $1$.  Finally, on lines 27-31
we backtrack through our lookup table $P$ that stores the optimal $(S-\{x\}, k)$
for a given $(S, x)$, appending each optimal $k$ value to the optimal tour $T$.

@@

\begin{proof}
  (Induction) For our basis step, consider that an empty set $\varnothing$ has $1$
  subset (itself).  $2^n = 2^0 = 1$, so a set with $0$ elements has $2^0$ subsets.\
  For induction, suppose a $k$-element set has $2^k$ subsets.  We will show every
  $(k+1)$-element subset has $2^(k+1)$ subsets.  Let $A=\{a_1,a_2,\dots,a_k, b\}$,
  notice it contains $k+1$ elements.  We partition the subsets of $A$ into two sets
  $B$ and $C$, where $B$ contains the subsets that contain $b$ and $C$ contains
  the subsets that don't contain $b$.  $C$ contains every subset of $A-{b}$, so it
  has $2^k$ elements.  By construction, it follows that $B$ must have the same number
  of elements as $C$, so it also has $2^k$ elements.  Since $B$ and $C$ are independent
  sets, the size of $B\union C$ is $2^k + 2^k = 2^{k+1}$.
\end{proof}

@@

\newpage
\appendix
\section*{Appendices}
\newpage
\addcontentsline{toc}{section}{Appendices}
\renewcommand{\thesubsection}{\Alph{subsection}}

\subsection{Collaboration}
  I collaborated with on this project Grant Eaton and Luke Artnak.
  \par

  Grant helped me search for primary sources to use in this paper.
  He also made a suggestion
  for space optimization on my Held-Karp implementation, which
  I agreed with, but did not implement.  I helped Grant
  with various Latex tasks and getting Python networkx code
  to work.

  Luke helped me analyze the runtime of the Held-Karp algorithm,
  which helped me greatly in writing the runtime proof.  He also
  made some suggestions on how I could make my Held-Karp python code
  more readable, which again, I did not follow through with.  I helped
  Luke get python networkx code working, although he ended up
  switching to JavaScript later on.

  \par

\newpage
\subsection{Source Code}
  Brute Force Implementation
  \inputminted[linenos=true]{python}{../src/brute_force.py}
  \newpage
  Held-Karp Implementation
  \inputminted[linenos=true]{python}{../src/held_karp.py}
  The output of the main program running on my machine.
  Yellow values are estimations.
  \includegraphics[scale=0.5]{screenshot.png}

\newpage
\subsection{Requirements}
For each of the requirements of the paper, I have given
the section where that requirement was met.
\begin{enumerate}
  \item Requirement: Describe the problem.  See sections 1.1, 1.2, and 1.3.
  \item Requirement: Explain your naive / BF implementation.  See section 2.
  \item Requirement: Motivate a solution that improves on the naive solution.
    See section 3.2 for better solution and 3.5 for a comparison.
  \item Requirement: Defend your solution against prior art: All of section
    3 contains cited work, primarily in section 3.5.
  \item Requirement: explain your improved solution by presenting pseudocode.  See section 3.6.
  \item Requirement: Provide a proof that your solution has some nontrivial property.  See section 3.7.
  \item Requirement: Experimentally compare your solution to BF / Naive approaches.  See section 4.
  \item Requirement: Do version A or B of the assignment.  I've done A by choosing the Held-Karp algorithm.
    Section 3.5 is explicitly states this paper is concerned with correctness, rather than speed.
\end{enumerate}
